Jay's Peer Assessment file 1
========================================================

Set working directory, download the data, unzip the files, read in the data:

```{r}
setwd("~/Documents/R/repdata")
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile="activity.zip", method="curl")
unzip("activity.zip")
activity <- read.csv("activity.csv")
```


summarize data
```{r tidydata}
library(plyr)
activity$date <- as.Date(activity$date)
# activity$interval <- strptime(activity$interval, "%H%M")
perDay <- ddply(activity, 
                ~date, 
                summarise, 
                stepsSum = sum(steps),
                stepsMean = mean(steps, na.rm =TRUE),
                stepsMedian = median(steps, na.rm =TRUE))
```

make a histogram of steps per day
```{r histogram}
library(ggplot2)
ggplot(perDay, aes(stepsSum)) +
geom_histogram(binwidth = 1600,
               colour = "blue",
               fill = "white") +
xlab("Steps Per Day")
```

return the mean and median
```{r meanMedian}
perDay
ggplot(activity, aes(date, steps)) +
     stat_summary(fun.y = mean, geom = "line")
```


plot time series
```{r timeseries}
intervals <- ddply(activity, 
                   ~interval, 
                   summarise,
                   intervalMean = mean(steps, na.rm = TRUE), 
                   intervalMedian = median(steps, na.rm = TRUE))
with(intervals, plot(intervalMean ~ interval, 
                     type = "l", 
                     ylab="Average steps taken",
                     xlab="time of day"))
```

the interval with the max mean steps per day:
```{r maxMean}
intervals$interval[intervals$intervalMean==max(intervals$intervalMean)]
```

total number of missing values
```{r missingValues}
missing <- is.na(activity$steps)
table(missing)
```

replace missing values with mean values for that interval
```{r replace missing}
replaceNA <- function(steps, interval) {
     if(!is.na(steps))
          return(steps)
     else 
          return(intervals$intervalMean[intervals$interval==interval])
}
activitynarm <- activity
for (i in 1:length(activitynarm$steps)) {
     activitynarm$steps[i] <- replaceNA(activitynarm$steps[i], activitynarm$interval[i])
}

# intervals2 is summary data with new, NA-removed data.
intervals2 <- ddply(activitynarm, 
                   ~interval, 
                   summarise,
                   intervalMean = mean(steps, na.rm = TRUE), 
                   intervalMedian = median(steps, na.rm = TRUE))
with(intervals2, plot(intervalMean ~ interval, 
                     type = "l", 
                     ylab="Average steps taken",
                     xlab="time of day"))
par(mfrow = c(1,2))
with(intervals2, plot(intervalMedian ~ interval, 
                     type = "l", 
                     main = "NAs removed",
                     ylab="Median steps taken",
                     xlab="time of day"))
with(intervals, plot(intervalMedian ~ interval, 
                     type = "l", 
                     main = "NA values preserved",
                     ylab="Median steps taken",
                     xlab="time of day"))
perDay2 <- ddply(activitynarm, 
                ~date, 
                summarise, 
                stepsSum2 = sum(steps),
                stepsMean2 = mean(steps, na.rm =TRUE),
                stepsMedian2 = median(steps, na.rm =TRUE))
perDay <- cbind(perDay, perDay2[,2:4])

par(mfrow = c(1,2))
with(perDay, hist(stepsSum2,
                  main = "NAs removed",
                  ylim = c(0,25),  
                  breaks = 15,
                  ylab="daily steps taken",
                  xlab="counts"))
hist(perDay$stepsSum, 
   #  yaxt = "n",
     ylim = c(0,25),
     breaks = 15,
     main = "NA values preserved",
     ylab = "daily steps taken",
     xlab = "counts")
               
```

Imputing missing values conclusion:
given the method I used, the mean values for the intervals should not change, however with the significant increase in non-zero values, the medians, or more precisely, the upper quartile should increase. Additionally, the total number of steps taken should significantly increase.


Now to compare average number of steps taken on weekdays versus weekends:

```{r weekdays}
# add days of the week to activitynarm data.
activitynarm$day <- weekdays(activitynarm$date)

# classify days as weekdays or weekends
for (i in 1:length(activitynarm$day)) {
     if (activitynarm$day[i] %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) {
          activitynarm$isWeekday[i] <- "Weekday"
          }
     else activitynarm$isWeekday[i] <- "Weekend"
}

intervals3 <- ddply(activitynarm, 
                   .(interval, isWeekday), 
                   summarise,
                   intervalMean = mean(steps, na.rm = TRUE), 
                   intervalMedian = median(steps, na.rm = TRUE))

library(lattice)
xyplot(intervalMean ~ interval | isWeekday, 
       type = "l",
       data = intervals3, 
       layout = c(1,2))
```